The algebraic hierarchy in type theory using type classes

### 2010-06-16 jar.tex
spelling

The ``native'' notion of equality in Coq is that of terms being convertible, which in this context constitutes Leibniz equality, naturally propositionalized by an inductive type family \lstinline|eq| with a single constructor \lstinline|eq_refl| of type \lstinline|Π (T: Type) (x: T), x ≡ x|, where ``\lstinline|a ≡ b|'' is notation for \lstinline|eq T a b|~\footnote{We diverge from Coq traditional and reserve ``\lstinline|=|'' for setoid equality, as this is the equality we will be working with most of the time.}. Importantly, since convertibility is a congruence, a proof of \lstinline|a ≡ b| lets us substitute \lstinline|b| for \lstinline|a| anywhere inside a term without further conditions.

@@

\begin{lstlisting}
  Definition reflexive {A: Type} (R: relation A): Prop := Π a, R a a.
\end{lstlisting}
(The curly brackets used for \lstinline|A| mark it as an implicit argument.)

@@

In the absence of type classes, these are all very real problems, and for this reason the two biggest formalizations of abstract algebraic structures in Coq today (namely CoRN\ref{corn} and Ssreflect\ref{ssreflect}), both use \emph{bundled} representation schemes, using records with one or more of the components as fields instead of parameters. For reflexive relations, the following is a fully bundled representation---the other end of the spectrum:
\begin{lstlisting}
  Record ReflexiveRelation: Type :=
    { rr_carrier: Type
    ; rr_rel: relation rr_carrier
    ; rr_proof: Π x, rr_rel x x }.
\end{lstlisting}
Superficially, this instantly solves the problems described above: reflexive relations can now be declared and passed as self-contained packages, and the \lstinline|rr_rel| projection now constitutes a ``canonical name'' for relations that are known to be reflexive, and we could bind a notation to it\footnote{While there is no conventional notation for reflexive relations, the situation is the same in the context of, say, a semiring, where we would bind ``$+$'' and ``$*$'' notations to the record field projections for the addition and multiplication operations, respectively.}.

@@

If we revert back to the predicate formulation of reflexive (et. al) relations, we \emph{could} still define \lstinline|EquivalenceRelation| in a bundled fashion without the need for equalities:
\begin{lstlisting}
  Record EquivalenceRelation: Type :=
    { er_carrier: Type
    ; er_rel: relation er_carrier
    ; er_refl: ReflexiveRelation er_carrier er_rel
    ; er_sym: SymmetricRelation er_carrier er_rel
    ; er_trans: TransitiveRelation er_trans er_rel }.
\end{lstlisting}
However, the same argument as before leads to the conclusion that \lstinline|EquivalenceRelation|, too, should be a predicate. Indeed, it would be rather strange for the interface of equivalence relations to differ qualitatively from the interface of reflexive relations.

@@

Since \lstinline|CommutativeMonoid| indirectly includes a \lstinline|SemiGroup| field, which in turn includes a \lstinline|Equivalence| field, having a \lstinline|SemiRing| proof means having two distinct proofs that the equality relation is an equivalence. This kind of redundant knowledge (which arises naturally in mathematics) is never a problem in our setup, because the use of operational type classes ensures that terms composed of algebraic operations and relations never refer to structure proofs. We find that this is a tremendous relief compared to approaches that \emph{do} intermix the two and where one must be careful to ensure that such terms refer to the ``right'' proofs of properties. There, even \emph{strong} proof irrelevance (which would make terms convertible that differ only in what proofs they refer) would not make these difficulties go away entirely, because high-level tactics that rely on quotation of terms require syntactic identity (rather than ``mere'' convertibility) to recognize identical subterms.

@@

The reader will note that we use names for properties like distributivity and absorption, and in fact these are type classes as well (which is why we declare them with \lstinline|:>|). It has been our experience that almost any generic property worth naming is worth representing as a type class, so that its proofs are be resolved as instances behind the scenes whenever possible. Doing this consistently minimizes administrative noise in the code, bringing us closer to ordinary mathematical vernacular. Indeed, we believe that type classes are an elegant and apt formalization of the seemingly casual manner in which ordinary mathematical presentation assumes implicit administration and use of a ``database'' of structural properties previously proved, much more so than existing solutions using, for instance, canonical structures.

@@

We feel that these issues are not limitations of the Coq system, but merely reflect the fact that notationally identifying these two different and interdependent maps is a typical example of an ``abus de notation'' that by its very nature is ill-suited to a formal development where software engineering concerns apply. Hence, we do not adopt this practice, and use ``\lstinline|fmap F|'' (name taken from the Haskell standard library) to refer to the arrow map of a functor \lstinline|F|.

>>>>

The ``native'' notion of equality in Coq is that of terms being convertible, which in this context constitutes Leibniz equality, naturally propositionalized by an inductive type family \lstinline|eq| with a single constructor \lstinline|eq_refl| of type \lstinline|Π (T: Type) (x: T), x ≡ x|, where ``\lstinline|a ≡ b|'' is notation for \lstinline|eq T a b|. Here  we diverge from Coq tradition and reserve ``\lstinline|=|'' for setoid equality, as this is the equality we will be working with most of the time. Importantly, since convertibility is a congruence, a proof of \lstinline|a ≡ b| lets us substitute \lstinline|b| for \lstinline|a| anywhere inside a term without further conditions.

@@

\begin{lstlisting}
  Definition reflexive {A: Type} (R: relation A): Prop := Π a, R a a.
\end{lstlisting}
The curly brackets used for \lstinline|A| mark it as an implicit argument.

@@

In the absence of type classes, these are all very real problems, and for this reason the two biggest formalizations of abstract algebraic structures in Coq today (namely CoRN\ref{corn} and Ssreflect\ref{ssreflect}), both use \emph{bundled} representation schemes, using records with one or more of the components as fields instead of parameters. For reflexive relations, the following is a fully bundled representation---the other end of the spectrum:
\begin{lstlisting}
  Record ReflexiveRelation: Type :=
    { rr_carrier: Type
    ; rr_rel: relation rr_carrier
    ; rr_proof: Π x, rr_rel x x }.
\end{lstlisting}
Superficially, this instantly solves the problems described above: reflexive relations can now be declared and passed as self-contained packages, and the \lstinline|rr_rel| projection now constitutes a ``canonical name'' for relations that are known to be reflexive, and we could bind a notation to it. While there is no conventional notation for reflexive relations, the situation is the same in the context of, say, a semiring, where we would bind ``$+$'' and ``$*$'' notations to the record field projections for the addition and multiplication operations, respectively.

@@

If we revert back to the predicate formulation of relations, we \emph{could} still define \lstinline|EquivalenceRelation| in a bundled fashion without the need for equalities:
\begin{lstlisting}
  Record EquivalenceRelation: Type :=
    { er_carrier: Type
    ; er_rel: relation er_carrier
    ; er_refl: ReflexiveRelation er_carrier er_rel
    ; er_sym: SymmetricRelation er_carrier er_rel
    ; er_trans: TransitiveRelation er_trans er_rel }.
\end{lstlisting}
However, the same argument as before leads to the conclusion that \mbox{\lstinline|EquivalenceRelation|,} too, should be a predicate. Indeed, it would be rather strange for the interface of equivalence relations to differ qualitatively from the interface of reflexive relations.

@@

Since \lstinline|CommutativeMonoid| indirectly includes a \lstinline|SemiGroup| field, which in turn includes a \lstinline|Equivalence| field, having a \lstinline|SemiRing| proof means having two distinct proofs that the equality relation is an equivalence. This kind of redundant knowledge (which arises naturally in mathematics) is never a problem in our setup, because the use of operational type classes ensures that terms composed of algebraic operations and relations never refer to structure proofs. We find that this is a tremendous relief compared to approaches that \emph{do} intermix the two and where one must be careful to ensure that such terms refer to the ``right'' proofs of properties. There, even \emph{strong} proof irrelevance (which would make terms convertible that differ only in what proofs they refer to) would not make these difficulties go away entirely, because high-level tactics that rely on quotation of terms require syntactic identity (rather than ``mere'' convertibility) to recognize identical subterms.

@@

The reader will note that we use names for properties like distributivity and absorption. Indeed, these are type classes as well (which is why we declare them with \lstinline|:>|). It has been our experience that almost any generic property worth naming is worth representing as a type class, so that its proofs will be resolved as instances behind the scenes whenever possible. Doing this consistently minimizes administrative noise in the code, bringing us closer to ordinary mathematical vernacular. Indeed, we believe that type classes are an elegant and apt formalization of the seemingly casual manner in which ordinary mathematical presentation assumes implicit administration and use of a ``database'' of structural properties previously proved, much more so than existing solutions using, for instance, canonical structures.

@@

We feel that these issues are not limitations of the Coq system, but merely reflect the fact that notationally identifying these two different and interdependent maps is an abuse of notation that by its very nature is ill-suited to a formal development where software engineering concerns apply. Hence, we do not adopt this practice, and use ``\lstinline|fmap F|'' (name taken from the Haskell standard library) to refer to the arrow map of a functor \lstinline|F|.
